Assignment: Course Project 1
========================================================


```{r}

library(dplyr)
library(ggplot2)
library(data.table)
library(car)

#Load and read file
activity  = read.csv("activity.csv")

```

## What is the mean of the total number of steps taken per day?

Sum steps for each day

```{r}
steps_by_day = rowsum(activity$steps, activity$date, na.rm = FALSE)
```

#### Histogram of the total steps taken each day


```{r}
hist(steps_by_day, main="Steps Taken Each Day", xlab="Total Steps")
```

Calculate the mean and median for steps taken each day

```{r}
steps_mean <- as.character(round(mean(steps_by_day, na.rm = TRUE),2))
steps_median <- as.character(round(median(steps_by_day, na.rm = TRUE),2))
```

The Mean steps each day is `r steps_mean`.  
The Median steps each day is `r steps_median`.

### What is the average daily activity pattern?

```{r}

steps_per_int <- rowsum(activity$steps,activity$interval, na.rm = TRUE)

# Convert interval to integer
activity$interval <- as.integer(activity$interval)

#get counts of each 5 minute intervals
count_int <- nrow(filter(activity, interval == max(interval), rm.na = TRUE))

# Mean of the total steps per interval
steps_per_int <- data.table(steps_per_int)
colnames(steps_per_int) <- c("steps")
steps_per_int$mean <- steps_per_int$steps/count_int

# Plot the timeseries graph
plot(ts(steps_per_int$mean), type='l', main='Average # of Steps Per 5 Minute Interval', ylab='Average Steps', xlab='5 Minute Intervals')
```

```{r}
#Retrieve the interval with the highest average steps

activity2 = select(activity,steps, interval)
activity2 = aggregate(steps~interval,activity2, mean, rm.na = TRUE)
max_5_min = filter(activity2, steps == max(steps) )
max_5_min_int = max_5_min$interval
```

The `r max_5_min_int` 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

## Imputing missing values

```{r}
# Get count of rows missing values
incomplete = activity[!complete.cases(activity),]
incomplete = nrow(incomplete)
```

The total number of rows with incomplete data is `r incomplete`

### Devise a strategy for filling in all of the missing values in the dataset. 
#### Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

complete_activity = activity

# Calculate the median of the mean per day
activity3 = select(activity,steps, date)
activity3 = aggregate(steps~date,activity3, mean, rm.na = TRUE)
med_of_meanpd = median(activity3$steps)

#Set steps with NA values to the m
complete_activity$steps[is.na(complete_activity$steps)] <- med_of_meanpd
```

### Make a histogram of the total number of steps taken each day

```{r}
adj_steps_by_day = rowsum(complete_activity$steps, complete_activity$date)

hist(adj_steps_by_day, main="Steps Taken Each Day", xlab="Total Steps")

```

### Calculate and report the mean and median total number of steps taken per day

```{r}
adj_steps_mean <- as.character(round(mean(adj_steps_by_day, na.rm = TRUE),2))
adj_steps_median <- as.character(round(median(adj_steps_by_day, na.rm = TRUE),2))

```

The Mean of the adjusted steps per day is `r adj_steps_mean`.   
The Median of the adjusted steps each day is `r adj_steps_median`.

This varies only slightly from the unadjusted mean and median steps per day. 

# Are there differences in activity patterns between weekdays and weekends?

```{r}
# create day of the week variable
complete_activity = mutate(complete_activity, day_of_week = weekdays(as.Date(date,'%Y-%m-%d')))
# change dayof_week to 'weekend' or 'weekday'
complete_activity$day_of_week = recode(complete_activity$day_of_week, "c('Monday','Tuesday', 'Wednesday', 'Thursday', 'Friday')='weekday'")
complete_activity$day_of_week <- recode(complete_activity$day_of_week, "c('Saturday','Sunday')= 'weekend'")

#aggregate by weekend and weekday
complete_activity2 = aggregate(steps~interval+day_of_week,complete_activity, mean, rm.na = TRUE)

# Make a panel plot containing a time series plot 
ggplot(complete_activity2, aes(x=interval, y=steps)) + geom_line() + facet_grid(day_of_week ~ .)

```

